---
title: "JLC-245 Project 4"
author: "Melanie Klein"
date: "`r Sys.Date()`"
output: html_document
---

Step 0: Load and install packages
```{r}
# install.packages('tidyverse')
# install.packages('explore')
```

```{r}
library(tidyverse)
library(explore)
```

Step 1: Get data
```{r}
setwd("/Users/melanie/Desktop/GitHub/Cities and Crime/Project 4")
hate <- read.csv("hate_crime.csv", stringsAsFactors = FALSE)
```

Step 2: Inspect data
```{r}
# Look at the column names
colnames(hate)
```

```{r}
# Look at the unique values for the agency type field
unique(hate$agency_type_name)
```

```{r}
# Look at the unique values for the area population groups
unique(hate$population_group_desc)
```

Step 3: Subset data
```{r}
# Make a subset of the data for just D.C.:
hate.dc <- subset(hate, hate$state_abbr == 'DC')
```

```{r}
# US
explore(hate)
```
- 80% offenders are white (40%) or unknown (40%)*
- #1 location name of bar or nightclub (more than 50%)
- victim types overwhelmingly are businesses*
- most of the time it's a single offense
- most of the time it's a single bias

- events per year, may want to consider
- adult victim count generally 1, juvenile generally 0, not very exciting
- ethnicity mostly not known, not very exciting
- offense name very messy, might be something
- bias description, may want to consider
- multiple offense and bias, not very exciting

Interesting fields, turn them into rankings to build the table
- events per year
- offender race
- offense name
- location name
- bias
- victim types
Don't account for all measures, but pull things that are uniquely DC and different from the norm

```{r}
# DC
explore(hate.dc)
```
- year is different than the national trend (steady increase), 2018 decrease
- race is very different, more than 50% black or african american
- location name very different, highway/road/alley/street/sidewalk more than 60% compared to nightclubs
- victim type very different, almost 90% individual instead of business
- offense and bias are in line with national

Interesting
- events per year, national is flat line, DC goes up
- offender race, 40% white US, 53% black in DC, completely different, look at black
- offense name, assault was the most national, DC it is intimidation almost 30%, also have destruction/damage/vandalism and aggravated assault as the other two that stand out, focus on intimidation because it is the highest
- location name, highway
- bias, antigay male
- victim types, individuals

Step 4: Filter data
```{r}
unique(hate.dc$agency_type_name)
```
only city and other for agency type name in DC

```{r}
# filter based on population
hate.similar <- hate %>% 
  filter(population_group_code == '1B') 
```

```{r}
# cities in population 1B
unique(hate.similar$pug_agency_name)
```

34 cities, including DC
Went from 280k incidents to 25k incidents.
Choose the 5 most similar to DC

Step 5: Summarize data
```{r}
# Create a new table summarizing events per area:
hate.events.sum <- hate.similar %>% 
  group_by(pug_agency_name) %>%
  summarise(count = n()) %>% 
  mutate(pct = round(count/sum(count)*100, 2))

hate.events.sum$diff <- abs(hate.events.sum$count - hate.events.sum$count[34])

events <- hate.events.sum[order(hate.events.sum$diff),]
events$rank.events <- seq.int(nrow(events)) - 1
```

```{r}
# offender race

#create summary table
hate.offender.race <- hate.similar %>%
  group_by(pug_agency_name, offender_race) %>%
  summarise(count = n()) %>%
  mutate(pct = round(count/sum(count)*100,2))

# filter by the field most unique to DC
offender.race <- subset(hate.offender.race,
                        hate.offender.race$offender_race == "Black or African American")

# calculate the difference between DC and the other cities
offender.race$diff <- abs(offender.race$pct - offender.race$pct[33])

# reorder the table
offender.race <- offender.race[order(offender.race$diff),]

# rank the table
offender.race$rank.race <- seq.int(nrow(offender.race)) - 1
```

```{r}
# offense type

#create summary table
hate.offense <- hate.similar %>%
  group_by(pug_agency_name, offense_name) %>%
  summarise(count = n()) %>%
  mutate(pct = round(count/sum(count)*100,2))

# filter by the field most unique to DC
offense <- subset(hate.offense,
                        hate.offense$offense_name == "Simple Assault")

# calculate the difference between DC and the other cities
offense$diff <- abs(offense$pct - offense$pct[34])

# reorder the table
offense <- offense[order(offense$diff),]

# rank the table
offense$rank.offense <- seq.int(nrow(offense)) - 1
```

```{r}
# location name

#create summary table
hate.location <- hate.similar %>%
  group_by(pug_agency_name, location_name) %>%
  summarise(count = n()) %>%
  mutate(pct = round(count/sum(count)*100,2))

# filter by the field most unique to DC
location <- subset(hate.location,
                        hate.location$location_name == "Highway/Road/Alley/Street/Sidewalk")

# calculate the difference between DC and the other cities
location$diff <- abs(location$pct - location$pct[34])

# reorder the table
location <- location[order(location$diff),]

# rank the table
location$rank.location <- seq.int(nrow(location)) - 1
```
```{r}
# bias

#create summary table
hate.bias <- hate.similar %>%
  group_by(pug_agency_name, bias_desc) %>%
  summarise(count = n()) %>%
  mutate(pct = round(count/sum(count)*100,2))

# filter by the field most unique to DC
bias <- subset(hate.bias,
                        hate.bias$bias_desc == "Anti-Gay (Male)")

# calculate the difference between DC and the other cities
bias$diff <- abs(bias$pct - bias$pct[32])

# reorder the table
bias <- bias[order(bias$diff),]

# rank the table
bias$rank.bias <- seq.int(nrow(bias)) - 1
```

```{r}
# victim types

#create summary table
hate.victim <- hate.similar %>%
  group_by(pug_agency_name, victim_types) %>%
  summarise(count = n()) %>%
  mutate(pct = round(count/sum(count)*100,2))

# filter by the field most unique to DC
victim <- subset(hate.victim,
                        hate.victim$victim_types == "Individual")

# calculate the difference between DC and the other cities
victim$diff <- abs(victim$pct - victim$pct[34])

# reorder the table
victim <- victim[order(victim$diff),]

# rank the table
victim$rank.victim <- seq.int(nrow(victim)) - 1
```

```{r}
# combine the 6 tables: events, offender.race, offense, location, boas, victim
names(events) <- c("City", "Count.Events", "PCT.Events", "Diff.Events", "Rank.Events")
names(offender.race) <- c("City", "Race", "Count.Race", "PCT.Race", "Diff.Race", "Rank.Race")
names(offense) <- c("City", "Offense", "Count.Offense", "PCT.Offense", "Diff.Offense", "Rank.Offense")
names(location) <- c("City", "Location", "Count.Location", "PCT.Location", "Diff.Location", "Rank.Location")
names(bias) <- c("City", "Bias", "Count.Bias", "PCT.Bias", "Diff.Bias", "Rank.Bias")
names(victim) <- c("City", "Victim", "Count.Victim", "PCT.Victim", "Diff.Victim", "Rank.Victim")
```

```{r}
# joins
hate.ranks <- events %>%
  left_join(offender.race, by = "City")
hate.ranks <- hate.ranks %>%
  left_join(offense, by = "City")
hate.ranks <- hate.ranks %>%
  left_join(location, by = "City")
hate.ranks <- hate.ranks %>%
  left_join(bias, by = "City")
hate.ranks <- hate.ranks %>%
  left_join(victim, by = "City")
```

```{r}
# fix column order
ranks <- hate.ranks[c(1, 5, 10, 15, 20, 25, 30)]
```

can create the threshold for similar
ranking will put you in the ballpark but isn't enough
could calculate an average ranking

Top 6 for each of six variables, not scientific but arbitrary
Seattle** 1 4 6 = 11
Portland** 3 3 6 = 12
Denver** 3 4 5 = 12
Atlanta** 2 4 5 = 11
Kansas City** 2 4 6 = 12

Columbus** 2 5 6 = 13
Detroit** 3 4 6 = 13

San Jose
Memphis
Milwaukee*
Albuquerque
Honolulu
Austin
Baltimore
San Francisco*
New Orleans
Louisville
Sacramento
Fort Worth
Tuscon

5 cities
- counties, agencies/jurisdictions
3 variables per similar city
- don't need 15 unique variables, can use as few as 3 but likely in the middle


# Data
The FBI's Hate Crime Statistics dataset spans from 1991 to 2023 and includes a total of 253,776 events reported at the incident level. It provides annual statistics on hate crimes, detailing offenses, victims, and offenders motivated by bias related to race, gender, gender identity, religion, disability, sexual orientation, or ethnicity. The data is collected from law enforcement agencies through both the Summary Reporting System (SRS) and National Incident-Based Reporting System (NIBRS) forms, contributing to the UCR Program. The dataset includes hate crime incidents reported across the United States, with data collected from law enforcement agencies at the................. (local, state, and federal) levels.

# Methods
2109 DC hate crimes

# Analysis
give numbers behind it, counts or percentages, don't just use rankings
use ranks to identify similarities
use hate.ranks for the graphs

Interesting fields, turn them into rankings to build the table
- events per year
- offender race
- offense name
- location name
- bias
- victim types

## City 1: Seattle

### Analysis: 
Seattle is similar to Washington, D.C. based on the attributes of number of events (rank 1 of 33), location name (rank 4 of 33), and victim type. Specifically, for number of events, 2,530 hate crimes have occurred in Seattle which is only 615 more than the 1,915 hate crimes that have occurred in Washington, D.C. Specifically, for location name, 39.21% of hate crimes in Seattle occurred on a highway/road/alley/street/sidewalk which is only 4.13% less than the 39.21% of hate crimes in Washington, D.C. that occurred on a highway/road/alley/street/sidewalk. Specifically, for victim type, 88.77% of hate crimes in Seattle were committed against an individual which is only 0.63% less than the 89.4% of hate crimes in Washington, D.C. that were committed against an individual.

- 1 events - done
- 4 location - done
- 6 victim - done

### Visual:
```{r}
ggplot(hate.ranks %>%
         mutate(highlight.seattle = ifelse(City %in% c("Seattle", "Washington"), "Highlighted", "Other")), 
       aes(y = reorder(City, -Rank.Events), x = Count.Events, fill = highlight.seattle)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Highlighted" = "red", "Other" = "steelblue"), guide = "none") +
  labs(title = "Number of Hate Crime Events by City",
       x = "Number of Events",
       y = "City (ordered by rank)") +
  theme_minimal()
```

## City 2: Portland

### Analysis: 
Portland is similar to Washington, D.C. based on the attributes of number of events (rank 3 of 33), location name (rank 6 of 33), and bias description (rank 3 of 31). Specifically, for number of events, 1,030 hate crimes have occurred in Portland which is only 885 fewer than the 1,915 hate crimes that have occurred in Washington, D.C. Specifically, for location name, 33.69% of hate crimes in Portland occurred on a highway/road/alley/street/sidewalk which is only 9.65% less than the 39.21% of hate crimes in Washington, D.C. that occurred on a highway/road/alley/street/sidewalk. Specifically, for bias description, 19.81% of hate crimes in Portland were due to anti-gay bias against a male which is only 15.7% less than the 35.51% of hate crimes in Washington, D.C. that were due to anti-gay bias against a male.

- 3 events - done
- 6 location - done
- 3 bias - done

### Visual:
```{r}
hate.ranks %>%
  filter(!is.na(Rank.Bias)) %>%
  mutate(highlight = ifelse(City %in% c("Washington", "Portland"), "Highlighted", "Other")) %>%
  arrange(Rank.Bias) %>%  # Order by Rank.Bias
  ggplot(aes(x = PCT.Bias, y = reorder(City, -Rank.Bias), color = highlight)) +
  geom_point(size = 3) +  # Plot the dots with size 3
  scale_color_manual(values = c("Highlighted" = "red", "Other" = "steelblue"), guide = "none") +
  labs(title = "Percent of Anti-Gay (Male) Bias by City",
       x = "Percent of Anti-Gay (Male) Bias",
       y = "City (ordered by rank)") +
  theme_minimal()
```

## City 3: Denver

### Analysis:
Denver is similar to Washington, D.C. based on the attributes of number of events (rank 4 of 33), victim type (rank 3 of 33), and location name (rank 5 of 33). Specifically, for number of events per year, 920 hate crimes have occurred in Denver which is only 995 fewer than the 1,915 hate crimes that have occurred in Washington, D.C. Specifically, for victim type, ... Specifically, for location name, 35.98% of hate crimes in Denver occurred on a highway/road/alley/street/sidewalk which is only 7.36% less than the 39.21% of hate crimes in Washington, D.C. that occurred on a highway/road/alley/street/sidewalk. Specifically, for victim type, 89.13% of hate crimes in Denver were committed against an individual which is only 0.27% less than the 89.4% of hate crimes in Washington, D.C. that were committed against an individual.

- 4 events - done
- 3 victim - done
- 5 location - done

### Visual: location
```{r}
ggplot(hate.ranks %>% 
         filter(!is.na(PCT.Location)),
       aes(x = "PCT.Location", y = reorder(City, PCT.Location), fill = PCT.Location)) + 
  geom_tile() +
  scale_fill_gradient(low = "steelblue", high = "red") +
  labs(title = "Heatmap of Location Name by City",
       x = "Percentage of Offenses (highway/road/alley/street/sidewalk)",
       y = "City") +
  geom_text(aes(label = scales::percent(PCT.Location / 100)), color = "white", size = 3) +  # Add percent labels
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

```

## City 4: Atlanta

### Analysis: 
Atlanta is similar to Washington, D.C. based on the attributes of offender race (rank 4 of 32), offense name (rank 2 of 33), and bias (rank 5 of 31). Specifically, for offender race, 35.48% of offenders in Atlanta are Black or African American which is only 14.75% less than the 50.23% of offenders in Washington, D.C. that are Black or African American. Specifically, for offense name, 32.36% of offenses in Atlanta are simple assault which is only 6.28% less than the 38.54% of offenses in Washington, D.C. that are simple assault. Specifically, for bias description, 19.35% of hate crimes in Atlanta were due to anti-gay bias against a male which is only 16.16% less than the 35.51% of hate crimes in Washington, D.C. that were due to anti-gay bias against a male.

- 4 race - done
- 2 offense - done
- 5 bias - done

### Visual:
```{r}
hate.ranks %>%
  filter(City %in% c("Washington", "Atlanta")) %>%
  mutate(PCT.Offense.Other = 100 - PCT.Offense) %>%  # Calculate the complement of PCT.Offense
  gather(key = "Category", value = "Percent", PCT.Offense, PCT.Offense.Other) %>%
  ggplot(aes(x = "", y = Percent, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +  # Use 'identity' to plot actual values
  coord_polar(theta = "y") +  # Convert to pie chart
  facet_wrap(~ City, strip.position = "bottom") +  # Move facet labels to the bottom
  scale_fill_manual(values = c("PCT.Offense" = "red", "PCT.Offense.Other" = "steelblue")) +  # Custom colors
  geom_text(aes(label = ifelse(Category == "PCT.Offense", paste0(round(Percent, 1), "%"), "")),  # Only show label for PCT.Offense
            position = position_stack(vjust = 0.5),  # Place labels at the center of the segments
            color = "white", size = 5) +  # White text for readability
  labs(title = "Percent of Simple Assault Offenses by City") +
  theme_void() +  # Remove background grid lines
  theme(legend.position = "none",  # Remove legend if not necessary
        strip.text = element_text(size = 12, hjust = 0.5))  # Adjust facet label size and position
```

## City 5: Kansas City

### Analysis: 
Kansas City is similar to Washington, D.C. based on the attributes of offender race (rank 6 of 32), offense name (rank 4 of 33), and victim type (rank 2 of 33). Specifically, for offender race, 30.83% of offenders in Atlanta are Black or African American which is only 19.4% less than the 50.23% of offenders in Washington, D.C. that are Black or African American. Specifically, for offense name, 30% of offenses in Kansas City are simple assault which is only 8.54% less than the 38.54% of offenses in Washington, D.C. that are simple assault. Specifically, for victim type, 89.17% of hate crimes in Kansas City were committed against an individual which is only 0.23% less than the 89.4% of hate crimes in Washington, D.C. that were committed against an individual.

- 6 race - done
- 4 offense - done
- 2 victim - done

### Visual:
```{r}
ggplot(hate.ranks %>%
         filter(City != "Washington") %>%
         mutate(highlight.kc = ifelse(City %in% c("Kansas City"), "Highlighted", "Other")), 
       aes(y = reorder(City, -Rank.Victim), x = Diff.Victim, fill = highlight.kc)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Highlighted" = "red", "Other" = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, max(hate.ranks$Diff.Victim), by = 1)) +
  labs(title = "Difference to D.C. of Individual Victim Type by City",
       x = "Absolute Percent Difference",
       y = "City (ordered by rank)") +
  theme_minimal()
```

Questions
- number of events? events per year?
- include ranks in analysis?
