---
title: "Project 1"
author: "Melanie Klein"
date: "`r Sys.Date()`"
output: html_document
---

Step 0: Load and install packages
```{r packages, message=FALSE, warning=FALSE, include=FALSE}
#install.packages('tidyverse')
#install.packages('explore')
#install.packages('GGally')
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(explore)
library(GGally)
```

Step 1: Examine data

Step 2: Read data test
```{r data, warning=FALSE}
dc.data2025 <- read.csv("https://opendata.arcgis.com/datasets/74d924ddc3374e3b977e6f002478cb9b_7.csv", stringsAsFactors = FALSE) 
dc.data2024 <- read.csv("https://opendata.arcgis.com/datasets/c5a9f33ffca546babbd91de1969e742d_6.csv", stringsAsFactors = FALSE)
dc.data2023 <- read.csv("https://opendata.arcgis.com/datasets/89561a4f02ba46cca3c42333425d1b87_5.csv", stringsAsFactors = FALSE)
dc.data2022 <- read.csv("https://opendata.arcgis.com/datasets/f9cc541fc8c04106a05a1a4f1e7e813c_4.csv", stringsAsFactors = FALSE)
dc.data2021 <- read.csv("https://opendata.arcgis.com/datasets/619c5bd17ca2411db0689bb0a211783c_3.csv", stringsAsFactors = FALSE)
dc.data2020 <- read.csv("https://opendata.arcgis.com/datasets/f516e0dd7b614b088ad781b0c4002331_2.csv", stringsAsFactors = FALSE)
dc.data2019 <- read.csv("https://opendata.arcgis.com/datasets/f08294e5286141c293e9202fcd3e8b57_1.csv", stringsAsFactors = FALSE)
dc.data2018 <- read.csv("https://opendata.arcgis.com/datasets/38ba41dd74354563bce28a359b59324e_0.csv", stringsAsFactors = FALSE)
dc.data2017 <- read.csv("https://opendata.arcgis.com/datasets/6af5cb8dc38e4bcbac8168b27ee104aa_38.csv", stringsAsFactors = FALSE)
dc.data2016 <- read.csv("https://opendata.arcgis.com/datasets/bda20763840448b58f8383bae800a843_26.csv", stringsAsFactors = FALSE)
dc.data2015 <- read.csv("https://opendata.arcgis.com/datasets/35034fcb3b36499c84c94c069ab1a966_27.csv", stringsAsFactors = FALSE)
dc.data2014 <- read.csv("https://opendata.arcgis.com/datasets/6eaf3e9713de44d3aa103622d51053b5_9.csv", stringsAsFactors = FALSE)
dc.data2013 <- read.csv("https://opendata.arcgis.com/datasets/5fa2e43557f7484d89aac9e1e76158c9_10.csv", stringsAsFactors = FALSE)
dc.data2012 <- read.csv("https://opendata.arcgis.com/datasets/010ac88c55b1409bb67c9270c8fc18b5_11.csv", stringsAsFactors = FALSE)
dc.data2011 <- read.csv("https://opendata.arcgis.com/datasets/9d5485ffae914c5f97047a7dd86e115b_35.csv", stringsAsFactors = FALSE)
dc.data2010 <- read.csv("https://opendata.arcgis.com/datasets/fdacfbdda7654e06a161352247d3a2f0_34.csv", stringsAsFactors = FALSE)
dc.data2009 <- read.csv("https://opendata.arcgis.com/datasets/73cd2f2858714cd1a7e2859f8e6e4de4_33.csv", stringsAsFactors = FALSE)
dc.data2008 <- read.csv("https://opendata.arcgis.com/datasets/180d56a1551c4e76ac2175e63dc0dce9_32.csv", stringsAsFactors = FALSE)
```

Step 3: Merge data
```{r merge, warning=FALSE}
data.temp <- rbind(dc.data2008, dc.data2009, dc.data2010, dc.data2011, dc.data2012, dc.data2013, dc.data2014, dc.data2015, dc.data2016, dc.data2017, dc.data2018, dc.data2019, dc.data2020, dc.data2021, dc.data2022, dc.data2023, dc.data2024, dc.data2025)
```

Step 4: Clean data
```{r clean, warning=FALSE}
dc.data <- separate(data.temp, REPORT_DAT, into = c("date", "time"), sep = " ")
```

Step 5: Create data
```{r warning=FALSE}
#Format the date column
dc.data$date <- as.Date(dc.data$date, format = "%Y/%m/%d")
#Create an hour of day column
dc.data$hour <- substr(dc.data$time, 0, 2)
dc.data$hour <- as.numeric(dc.data$hour)
#Create day of week
dc.data$dow <- weekdays(dc.data$date)
#Create week of year
dc.data$week <- format(as.Date(dc.data$date), "%U")
dc.data$week <- as.numeric(dc.data$week)
#Create month of year
dc.data$month <- months(dc.data$date)
#Create year
dc.data$year <- substr(dc.data$date, 0, 4)
```

Step 6: Analyze data
```{r explore, warning=FALSE}
explore(dc.data)
```

Step 7: Visualize data
```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
data.new <- select(dc.data, hour, DISTRICT, WARD, week, year)
ggpairs(data.new, cardinality_threshold = 18) + theme_bw()
```

Step 8: Intro to graphing data
```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
ggplot(dc.data, aes(SHIFT)) + geom_bar(stat = "count", aes(fill = OFFENSE), position = position_stack(reverse = TRUE)) + theme_classic() + facet_wrap(~DISTRICT)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(dc.data, aes(SHIFT)) + geom_bar(stat = "count", aes(fill = OFFENSE), position = position_stack(reverse = TRUE)) + theme_classic() +
  labs(
    title = "Your title here",
    x = "Shift",
    y = "y axis label here"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(),
    axis.ticks = element_blank()
  )
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
dc.data$TYPE <- case_when(
  dc.data$OFFENSE %in%
    c(
      "ARSON",
      "BURGLARY",
      "MOTOR VEHICLE THEFT",
      "THEFT F/AUTO",
      "THEFT/OTHER") ~ "Property",
  dc.data$OFFENSE %in%
    c("ASSAULT W/DANGEROUS WEAPON", "HOMICIDE", "ROBBERY", "SEX ABUSE")
  ~ "Person")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
dc.data %>% ggplot() +
  geom_line(aes(x = hour), stat = "count", group = 1, color = "blue", size = 1) +
  labs(title = "DC Crimes by Hour Reported \n2008 - 2025", x = "Hour Reported", y = "Number of Crimes (Thousands)", fill = "Type of Crime") +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        axis.text.x = element_text(color = "black", size = 10),
        axis.text.y = element_text(color = "black", size = 10),
        axis.ticks.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "gray"),
        panel.background = element_blank()) +
  scale_y_continuous(limits = c(0, 50000), breaks = c(0, 5000,10000,15000,20000,25000,30000,35000), labels = c(0,5,10,15,20,25,30,35))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
dc.data %>% filter(DISTRICT != "NA") %>%
  ggplot() +
  geom_bar(aes(x = as.factor(DISTRICT), fill = TYPE), stat = "count") +
  labs(title = "Crimes by DC Police District \n2010 - 202x", x = "District", y = "Number of Crimes (Thousands)", fill = "Type of Crime") +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        axis.text.x = element_text(color = "black", size = 10),
        axis.text.y = element_text(color = "black", size = 10),
        axis.ticks.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "gray"),
        panel.background = element_blank()) +
  scale_y_continuous(limits = c(0, 125000), breaks = c(0, 25000,50000,75000,100000,125000), labels = c(0,25,50,75,100,125))
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
dc.data %>% filter(TYPE == "Person") %>%
  ggplot() +
  geom_bar(aes(x = OFFENSE, fill = METHOD), position = "fill") +
  labs(title = "DC Person Crimes by Method \n2010 - 2025", x = "Crime", y = "Proportion of Crime", fill = "Method") +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        axis.text.x = element_text(color = "black", size = 10),
        axis.text.y = element_text(color = "black", size = 10),
        axis.ticks.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "gray"),
        panel.background = element_blank()) +
  scale_x_discrete(labels = c("Assault", "Homicide", "Robbery", "Sex Abuse")) +
  scale_y_continuous(labels = c("0%", "25%", "50%", "75%", "100%")) +
  scale_fill_manual(values = c("#1b9e77", "#d95f02","#7570b3"), labels = c("Gun", "Knife", "Other"))
```



robbery occurs disproportionately at midnight in DC
homicides only have been logged as occurring at midnight in 17 years, could be a data integrity issue

shift. crimes occur the midnight shift 20% of time. evening and day occur nearly twice. breakdown is disproportionate. crime occurs disproportionately across shifts 

day of week by shift. saturday and sunday midnight shift. weekdays day and evening